Chaotic dispersal of tidal debris

### 2015-09-21 ms.tex
andreas comments

Several long, dynamically cold stellar streams have been observed around the Milky Way Galaxy, presumably formed from the tidal disruption of globular clusters.
In integrable potentials---where all orbits are dynamically regular---tidal debris phase-mixes close to the orbit of the progenitor system. 
However, cosmological simulations of structure formation suggest that the Milky Way's dark matter halo is expected not to be fully integrable; an appreciable fraction of orbits will be chaotic.
This paper examines the influence of chaos on the phase-space morphology of cold tidal streams.
We find very stark results:
Streams in chaotic regions look very different from those in regular regions.
We find that streams (simulated using test particle ensembles of nearby orbits) can be sensitive to chaos on a much shorter time-scale than any standard prediction (from the Lyapunov or frequency-diffusion times).
For example, on a weakly chaotic orbit with a chaotic timescale predicted to be $>$1000 orbital periods ($>$1000 Gyr), the resulting stellar stream is, after just a few 10's of orbits, substantially more diffuse than any formed on a nearby but regular orbit.
We find that the enhanced diffusion of the stream stars can be understood by looking at the variance in orbital frequencies of orbit ensembles centered around the parent (progenitor) orbit.
Our results suggest that long, cold streams around our Galaxy must exist only on regular (or very nearly regular) orbits; they potentially provide a map of the regular regions of the Milky Way potential.
This suggests a promising new direction for the use of tidal streams to constrain the distribution of dark matter around our Galaxy.

@@

The morphological (density) evolution of the tidal debris therefore depends on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also determined by the shape and radial profile of the gravitational potential of the host galaxy. By modeling the observed phase-space density of stream stars along with the host galaxy potential it is hoped that we may infer the 3D mass distribution of the host. Many tidal streams are observed around the Milky Way, M31, and other nearby galaxies; the known streams span a large range of distances---from $\approx10$ to $100~{\rm kpc}$---and progenitor masses---from $\approx$10$^3$ to $\approx$10$^8~\msun$ in stellar mass---\citep[][]{ibata94,odenkirchen01,belokurov06,grillmair06a,grillmair06b,bonaca12}. There has been extensive work on developing methods to use data from these streams to measure properties of Milky Way's dark matter halo. These methods span a range of complexity from orbit-fitting, to \emph{Streakline} \citep{kuepper12} or `particle-spray' models \citep{gibbons14}, to action-space density modeling \citep[e.g.,][]{sanders14, bovy14}, to N-body simulations \citep[e.g.,][]{law10}. All methods have been tested in some way on simulated observations of data and these tests typically demonstrate the recovery of parameters for analytic, static potential forms.

@@

We have only considered low-mass progenitor systems such as globular clusters because the intrinsic spreads in fundamental frequencies are small (0.1--0.5\%). Small changes to the frequencies of the orbits of tidal debris disrupted from these progenitors due to chaotic processes will therefore cause observable changes to the real-space morphology of the debris. For more massive progenitor systems, the typical size and velocity dispersion of the debris will be larger and thus the debris morphology will be less sensitive to small changes in orbital properties. The mass-scale of the debris that will display enhanced density evolution depends on the magnitude of weak chaos, which depends on the orbit structure of a given potential. In potentials with more significant chaos, debris disrupted from more massive progenitor systems may also display `stream-fanning.' However, since the scale of the debris is larger it is more likely that some orbits will exist in chaotic regions.

@@

Tidal stream modeling is one of the most promising ways to constrain the 3D mass distribution around the Milky Way at distances of 10s to 100s of kpc. Methods that use tidal streams to infer properties of the Galactic potential typically operate by constructing models of the debris distribution using either the present-day phase-space density \citep[e.g.,][]{kuepper12, kuepper15, gibbons14}, time-of-disruption phase-space density \citep{apw13, apw14}, or the density in angle-action coordinates \citep{sanders14, bovy14}. All of these methods may fail or produce uninterpretable results if modeling globular-cluster streams on mildly chaotic orbits. For each of these methods, it is important to understand the failures and biases introduced by ignoring chaotic orbit evolution.

@@

The Ophiuchus stream \citep{bernard14, sesar15} appears to be a thin, short tidal stream (deprojected length $\approx$1.5--2 kpc) near the Galactic bulge with no apparent progenitor. At Galactocentric $R \approx 1$ kpc, $z \approx 5$ kpc, the orbits of the stream stars likely pass through the MW disk, feel the triaxiality of the Galactic bar \citep[e.g.,][]{wegg13, wegg15}, and have short orbital periods (relative to streams in the halo). It is possible that the observed debris is the last remnants of the recently disrupted progenitor system \citep{sesar15}, however if a significant number of stars were disrupted on previous pericentric passages, this older debris may be `fanned' and still near the observed portion of the stream. The fanned debris would be much lower surface brightness and thus much more difficult to detect. The detection of this low-surface-brightness component would open up the possibility that enhanced density evolution due to chaos is a dynamically relevant process for thin streams in real galaxies, though \cite{carlberg15} has shown that it is also possible to get short, high density segments of streams from debris formed on eccentric orbits.

>>>>

Several long, dynamically cold stellar streams have been observed around the Milky Way Galaxy, presumably formed from the tidal disruption of globular clusters.
In integrable potentials---where all orbits are dynamically regular---tidal debris phase-mixes close to the orbit of the progenitor system. 
\chchchanges{However, the Milky Way's dark matter halo is expected not to be fully integrable; an appreciable fraction of orbits will be chaotic.}
This paper examines the influence of chaos on the phase-space morphology of cold tidal streams.
\chchchanges{Streams in chaotic regions look very different from those in regular regions.}
\chchchanges{We find that streams can be sensitive to chaos on a much shorter time-scale than any standard prediction (from the Lyapunov or frequency-diffusion times).}
For example, on a weakly chaotic orbit with a chaotic timescale predicted to be $>$1000 orbital periods ($>$1000 Gyr), the resulting stellar stream is, after just a few 10's of orbits, substantially more diffuse than any formed on a nearby but regular orbit.
We find that the enhanced diffusion of the stream stars can be understood by looking at the variance in orbital frequencies of orbit ensembles centered around the parent (progenitor) orbit.
Our results suggest that long, cold streams around our Galaxy must exist only on regular (or very nearly regular) orbits; they potentially provide a map of the regular regions of the Milky Way potential.
This suggests a promising new direction for the use of tidal streams to constrain the distribution of dark matter around our Galaxy.

@@

The morphological (density) evolution of the tidal debris therefore depends on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also determined by the shape and radial profile of the gravitational potential of the host galaxy. By modeling the observed phase-space density of stream stars along with the host galaxy potential it is hoped that we may infer the 3D mass distribution of the host. Many tidal streams are observed around the Milky Way, M31, and other nearby galaxies; the known streams span a large range of distances---from $\approx10$ to $100~{\rm kpc}$---and progenitor masses---from $\approx$10$^3$ to $\approx$10$^8~\msun$ in stellar mass---\citep[][]{ibata94,odenkirchen01,belokurov06,grillmair06a,grillmair06b,bonaca12}. There has been extensive work on developing methods to use data from these streams to measure properties of \chchchanges{the} Milky Way's dark matter halo. These methods span a range of complexity from orbit-fitting, to \emph{Streakline} \citep{kuepper12} or `particle-spray' models \citep{gibbons14}, to action-space density modeling \citep[e.g.,][]{sanders14, bovy14}, to N-body simulations \citep[e.g.,][]{law10}. All methods have been tested in some way on simulated observations of data and these tests typically demonstrate the recovery of parameters for analytic, static potential forms.

@@

We have only considered low-mass progenitor systems such as globular clusters because the intrinsic spreads in fundamental frequencies are small (0.1--0.5\%). Small changes to the frequencies of the orbits of tidal debris \chchchanges{stripped} from these progenitors due to chaotic processes will therefore cause observable changes to the real-space morphology of the debris. For more massive progenitor systems, the typical size and velocity dispersion of the debris will be larger and thus the debris morphology will be less sensitive to small changes in orbital properties. The mass-scale of the debris that will display enhanced density evolution depends on the magnitude of weak chaos, which depends on the orbit structure of a given potential. In potentials with more significant chaos, debris disrupted from more massive progenitor systems may also display `stream-fanning.' However, since the scale of the debris is larger it is more likely that some orbits will exist in chaotic regions.

@@

Tidal stream modeling is one of the most promising ways to constrain the 3D mass distribution around the Milky Way at distances of 10s to 100s of kpc. Methods that use tidal streams to infer properties of the Galactic potential typically operate by constructing models of the debris distribution using either the present-day phase-space density \citep[e.g.,][]{varghese11, kuepper12, kuepper15, gibbons14}, time-of-disruption phase-space density \citep{apw13, apw14}, or the density in angle-action coordinates \citep{sanders14, bovy14}. All of these methods may fail or produce uninterpretable results if modeling globular-cluster streams on mildly chaotic orbits. For each of these methods, it is important to understand the failures and biases introduced by ignoring chaotic orbit evolution.

@@

The Ophiuchus stream \citep{bernard14, sesar15} appears to be a thin, short tidal stream (deprojected length $\approx$1.5--2 kpc) near the Galactic bulge with no apparent progenitor. At Galactocentric $R \approx 1$ kpc, $z \approx 5$ kpc, the orbits of the stream stars likely pass through the MW disk, feel the triaxiality of the Galactic bar \citep[e.g.,][]{wegg13, wegg15}, and have short orbital periods (relative to streams in the halo). It is possible that the observed debris is the last remnants of the recently disrupted progenitor system \citep{sesar15}, however if a significant number of stars were \chchchanges{stripped} on previous pericentric passages, this older debris may be `fanned' and still near the observed portion of the stream. \chchchanges{The fanned debris would have significantly lower surface brightness and thus would be more difficult to detect.} The detection of this low-surface-brightness component would open up the possibility that enhanced density evolution due to chaos is a dynamically relevant process for thin streams in real galaxies, though \cite{carlberg15} has shown that it is also possible to get short, high density segments of streams from debris formed on eccentric orbits.
